Complementary roles of Theta and Alpha in Multi-Item Associations

### 2019-05-20 diff/manuscript_old.tex
adding diff

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declaritive memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically, oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding and retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the information that is encoded into the underlying representation.

@@

It is generally accepted that long-term memory encoding and retrieval are not isolated functions of any one brain area \cite{Gabrieli98,PrinceDaselaarCabeza05}.  However, there are many challenges in moving beyond the traditional studies focusing on individual regions  (e.g., concentrating on the prominent role of the hippocampus), to obtain insight into the nature of network-level dynamics.  An increasing body of work support the idea that oscillatory dynamics in multiple brain regions are associated with different aspects of  memory encoding and retrieval \cite{NyhusCurran10,HanslmayrStaudiglFellner12,Klimesch12}, %update refs here!!!
These studies have generally found a consistent relationship between several frequency bands and successful encoding or retrieval of experiences; namely that oscillatory power within the alpha (8 to 12Hz) and beta (13 to 30Hz) frequency bands generally decrease with the successful execution of these memory processes, while theta (3 to 8Hz) and gamma (30 to 100+Hz) increase in the regions directly involved in the memory operations. Why these relationships exist, and what they tell us about the larger functional organization of memory systems in the brain are still open questions.

One model suggests that thalamic circuits play a functional role in selectively gating communication between cortical regions by synchronizing them in various frequency bands \cite{ShermanGuillery06,SaalmannKastner11,KetzJensenOReilly15}.  These gating mechanisms, implemented  by oscillatory synchronization, are central to system level brain function and ultimately affect cognitive behavior such as memory encoding and retrieval.  Further, the particular oscillatory dynamics engaged by this selection process may provide indicators of which system-scale neural circuits are coordinating together to facilitate that particular form of memory processing.

A particular thalmo-cortical circuit of interest for memory processes involves the hippocampus and its associated subcortical network, including the anterior thalamus, and medial septum which are essential in the maintenance of theta oscillations (4 to 8 Hz) \cite{Buzsaki02}.  Theta oscillations have been shown to be critically involved in successful encoding and retrieval of episodic like memories in both humans and animals \cite{NyhusCurran10,JonesWilson05,FuentemillaBarnesDuzelEtAl14}, and are believed to be primarily driven by the pacemaker cells within the medial septum.  The anterior thalamus, however,  provides the connectivity for the hippocampus to potentially modulate communication with the medial and orbital PFC through theta synchronization \cite{VanderWerfJollesWitterEtAl03,AggletonDumontWarburton11}.  Thus, the theta circuit is important for the core functioning of the hippocampus during encoding and retrieval.

@@

To test this inverse relationship between theta and alpha we choose a recent hippocampal dependent multi-item association study to base our computational and empirical investigations on \cite{HornerBisbyBushEtAl15} (see Figure \ref{fig:deev}).  In this study the authors used a previously developed paradigm (see \textcite{HornerBurgess13,HornerBurgess14}) in which they manipulated the binding of a set of overlapping paired associates.   Here two types of associative stimuli were studied, closed-loop and open-loop. In the `closed-loop' events participants studied a set of paired associates on separate trials, e.g. Person-Location, Location-Object, and Object-Person, where each category refers to the same stimulus for a given event.  This created a complete set of binding between all elements.   In contrast, the `open-loop' events incorporated a new element and left some relationships unstudied, e.g. Object-Location, Location-Person, Person-Animal.  Each of the associations was later tested in a cued recognition task, which showed greater inter-event Dependency for the closed-loop compared to the open-loop associates.  Dependency here is operationalized as the retrieval of \emph{all} associated pairs from a given event, or none; whereas no Dependency would imply retrieval of some of the associated pairs but not all.  The goal of this metric is to capture the strength of the binding between elements within a given event, where high Dependency indicates strong associations leading to retrieval of all items within a given event, and low Dependency indicates weak associations leading to some of the items being retrieved incorrectly. In contrast, a non-associative measure of accuracy can be made using an Independent model which calculates Dependency based on the assumption that each retrieval trial is statistically independent from the next, see Section \ref{sec:CalcDep} for more details.  %This provides a nice measure of both Recollection yielding high Dependency from strong associations between elements, but also uses a recognition paradigm where successful retrieval with low between element Dependency implies a similar construct to Familiarity, i.e. retrieval without detailed recall.

@@

The other critical aspect of this bottom-up attention mechanism is the connection to alpha oscillations.  Historically, alpha power has been shown to correlate with the onset of sleep, which led to the suggestion that high alpha power indicates an idle or perhaps inhibited cortical region.  The mechanistic connection between alpha power and thalamic gating is captured in the firing properties of the thalamic relay cells (TRC) which show two distinct modes of firing.  The first shows oscillatory firing patterns acting like a pacemaker system in the alpha frequency range (approximately 10 Hz or 100ms periodicity), and is generally referred to as bursting.  This bursting mode is believed to be engage when a given cortical region is in an idle state, i.e. when no active processing is occurring in the thalamus or the corresponding cortex linked to that thalamic relay region.  The other regime, referred to as tonic firing, shows no specific oscillatory firing pattern, and instead shows increased spontaneous firing in both the thalamus and corresponding cortical regions \cite{LerescheLightowlerSolteszEtAl91,ShermanGuillery06,LopesdaSilva91}.  These dynamics have been studied extensively as a naturally occurring phenomena of thalamic interactions with cortex, however simulation studies have also shown the potential for these mechanisms to be used in a goal directed way \cite{HindriksvanPutten13,VijayanKopell12,LeeWhittingtonKopell13}. The implications of these results are that cortical regions being driving by TRCs in bust mode will show more alpha power, and will process the incoming cortical-cortical information less veridically, while cortical regions being driven by tonic TRC firing will show a relative decrease in alpha power and process cortical-cortical information more completely.  This is the foundation for the connection between alpha power and thalamic gating, such that increased alpha power is indicative of decreased cortical processing.

@@

In this work we start with the Theta-Phase model of the hippocampus and entorhinal cortex, as described in \textcite{KetzMorkondaOReilly13} and implemented in the Emergent framework \cite{AisaMingusOReilly08} (see Figure \ref{fig:depNet}).  Briefly, this model is built upon a series of structural and functional hypotheses based on anatomical and physiological data, which have been captured in the complementary learning systems (CLS) model of the hippocampus \cite{McClellandMcNaughtonOReilly95, NormanOReilly03}. The Entorhinal Cortex (EC) in the model is assumed to be the cortical gateway to the hippocampus.  This gateway feeds through the tri-synaptic pathway to the Dentate Gyrus (DG), CA3, and then to CA1.  Similarly, there is a parallel connection through the mono-synaptic pathway from the EC to the CA1 (and back).

@@

The learning dynamics within this model, as governed by the LEABRA algorithm, are dictated by the phases of a presumed theta oscillation, thus its short-hand name of 'Theta-Phase' used here. In short, during learning two error signals are derived within every period of a give theta oscillation.  The first, during the trough of the theta oscillation, provides an error driven learning signal to the mono-synaptic pathway, learning the invertible mapping between EC and CA1 can be thought of as an encoding.  The second, during the peak of the theta oscillation, provides a error driven learning signal on the tri-synaptic pathway, learning to recollect missing or incomplete information can be thought of as retrieval.

@@

\subsection{Testing Procedure}
\label{sec:TestProc}
The encoding phase was split into three mini-blocks. One pairwise association for each of the 36 events was presented during each of the three mini-blocks. Presentation order within each mini-block was randomized. For the closed-loop events, the order of presentation across the three mini-blocks for a specific event was either: (1) location-object/animal, location-person and person-object/animal, (2) person-object/animal, location-person and location-object/animal, (3) location-person, location-object/ animal, person-object/animal or (4) location-person, person-object/animal and location-object/animal. For open-loop events, it was either: (1) location-object, location-person, person-animal, (2) person-animal, location-person, location-object, (3) location-person, location-object, person-animal or (4) location-person, person-animal, location-object. Each of the four orderings was run on two random initialization of weights, to make up the pool of results. The model was trained on 2 epochs (random permutations of all the trials within a mini-block) for each mini-block before starting training on the next mini-block.  Thus, for the first and second mini-blocks, the closed-loop and open-loop conditions were identical, presenting a single pairwise association and then a second overlapping pairwise association. In contrast, the third mini-block either formed an all-to-all associative structure in the closed-loop condition or an associative chain in the open-loop condition as shown in Figure \ref{fig:deev}.

At retrieval, the model was tested on each encoded pairwise association for each event in both directions.  In this way, each event was tested across six separate retrieval trials.  In the behavioral paradigm each retrieval trial was a six-forced choice recognition test with a particular cue element type and potential 6 targets all from the same element type (for example, cue location, retrieve the associated person among five people from other events).  However, in the model, this was captured by providing an activation of 0.1429 to all units in the target element dimension, providing a similar net activation compared to the complete patterns (i.e. $49 * 0.1429=7$), and thus opening the thalamic gate to allow processing in the target dimension. Note that a single cue pattern, and non-informative target dimension activation were presented on all retrieval trials with the task being to reactivate the original paired associate.  This reactivation success was measured by the binary Name Error metric described in Section \ref{sec:HippModelValidation}, i.e. either correct reactivation or not for a given cue-target pair. Thus, as with encoding, the closed-loop and open-loop conditions were exactly matched in terms of stimuli and task demands, differing only in the potential occurrence of incidental reactivation (that is, different levels of pattern completion).  
%An example of what this looks like in the network is shown in Figure \ref{fig:netNonTarg}B, with more pattern completion occurring in the location stimulus dimension (both in ECout and the TRC layer) for the closed-loop condition compared to open-loop.

@@

The start of each cued recognition trial showed a fixation cross with 0.5 to 1.5 s uniform random duration, followed by the presentation of the cue word above an array of 6 target words (stacked 3 horizontally by 2 vertically, numerically label from left to right and top to bottom), and the fixation cross between them.  The cue word and one of the target words were randomly drawn from a particular event, and the other 5 target words were drawn from the same element type as the correct target word, but randomly from other events.  Locations of the target words in the array were determined randomly.  Participants were instructed to press the numeric key corresponding to the target word that was previously studied with the cue word.  After a button press was a made, a rectangle appeared on-screen for 0.5 s, highlighting the selected item.  Participants were not made aware of the underlying event structure, or the open and closed-loop conditions.

@@

All EEG data was taken from retrieval trials where all processing steps and analyses were done in MATLAB using in-house scripts (mat-mvm, n.d.) and the FieldTrip toolbox \cite{OostenveldFriesMarisEtAl11}. A high-pass filter at 0.1 Hz, low-pass filter at 100 Hz, and a notch filter from 59--61 Hz were applied to the data.  ICA artifact removal targeting blinks, and movement artifacts was done on the continuous (i.e. pre-epoched) EEG data after removal of large artifacts through coarse visual inspection.  A semi-automated approach to artifact estimation and detection was done using the SASICA plugin ported for FieldTrip data structures, and methods described in \textcite{ChaumonBishopBusch15}.  After ICA artifact removal, each cued recognition trial was epoched in two ways: stimulus locked (SL) and response locked (RL).  SL epochs were 5000 ms segments, 1 s before the onset of the cue and target array, and 4 s after. RL epochs were also 5000 ms segements, 4 s before target response was made and 1 s after.  Fieldtrips's automatic artifact detection was used to reject particularly noisy epochs as determined by a $z$-score threshold of 20, as well as those that exceed an amplitude of $\pm$200 $\mu$V or a maximum change in amplitude of 200 $\mu$V.

@@

EEG results were analyzed according to the difference between open and closed-loop conditions.  To achieve this, $time \times electrode$ clusters were determined across subjects for each a-priori defined frequency band using a cluster based permutation test \cite{MarisOostenveld07}.  Frequency data were first averaged over a the defined frequency bands: 3-8 Hz for theta, 8-12 Hz for alpha and 12-30 Hz beta.  These data were then analyzed across all electrodes except the 4 surrounding the eyes, and across the time window of 500 to 3000 ms relative to the cue-target array onset in the stimulus locked epochs, and -3000 ms to -500 ms relative to the button press for a given trial in response locked epochs.  Clustering was done by performing a $t$-test comparing open and closed-loop conditions within each time/electrode bin across subjects for each of the target frequency ranges (theta, alpha and beta).  These $t$-values were then thresholded at $p<.05$, and clustered together based on spatial and temporal adjacency.   Cluster significance was calculated using a monte-carlo style permutation test of the summed $t$-values within a given cluster.  Each observed cluster was subject to 500 random permutations of condition labels where its significance was estimated by the proportion of random permutations which yielded clusters that had a summed $t$-value as large or larger than the observed cluster.

\subsubsection{Representation Similarity Analysis}
\label{sec:deevRSA}
The Representation Similarity Analysis (RSA), was motivated by the non-target reactivation seen in \textcite{HornerBisbyBushEtAl15} and was based on the methods originally developed by \textcite{KriegeskorteMurBandettini08}.  The underlying structure used in the RSA was based on individual retrieval trials within a given event and was done in two ways.  The first was the standard correlation approach where a similarity matrix was constructed by correlating the full $time \times frequency \times electrode$ matrix, splayed out to a single vector, pairwise with each retrieval trial within a given event.  Ideally this results in a 6x6 matrix with all 6 retrieval trials per event correlated with each other, however some retrieval trials were removed due to the presence of artifacts and therefore all events with more then 3 intact retrieval trials were used in the RSA.  The off diagonal $R$-values in this matrix were then converted to $z$-values using the inverse hyperbolic tangent, and averaged to achieve a single `similarity' measure across retrieval trials for a given event.  These were then averaged across condition with the prediction that more non-target reactivation would manifest in a higher similarity value averaged across events within a given condition.

@@

\begin{figure}
    \begin{center}
	\includegraphics[width=\textwidth]{figs/depNonTarg_edit.pdf}
    \end{center}
      \caption[Non Target Reactivation]{
       \emph{(A)} Name Error accuracy for non-target reactivation in open-loop and closed-loop events.  The horizontal axis labels correspond to the cue and target respectively, and the non-target element is the one not listed in the label.  For example `loc-per' indicates a location cue and person target trial, and therefore the non-target element is either object or animal depending on the encoding paris for the given event.  The 'oth' label corresponds to either the object or animal element again depending on the corresponding encoding pairs for the given event.  Note, non-target accuracy was not assessed for object or animal elements as cues, as open-loop events have no corresponding non-target element in these cases, and would be a biased comparison with closed-loop events which do. Black dots indicate individual network runs with different training orders (see section \ref{sec:TestProc}), red bars are closed-loop and blue bars are open-loop, dash lines are means across all cue-targ trial types, and error bars are 95\% confidence intervals. \emph{(B)} Example cued retrieval trials from open-loop (top) and closed-loop (bottom) events.  Both show a retrieval trial with a person cue (unit group 2, as numbered to the left of the Input layers), and a animal target (unit group 4), and a location non-target (unit group 1).  TRC shows bottom-up (person and animal unit groups) as well as hippocampal driven (Location unit group) modulation of thalamic gate.
	}
     \label{fig:netNonTarg}
\end{figure}

@@

\subsection{Behavioral Results}\label{sec:eegresults}

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{figs/deevEEGBeh_edit.pdf}
	\end{center}
	\caption[EEG Dependent Events Behavior]{
		\emph{(A)} Cued recognition accuracy for all cue-targ and targ-cue combinations, averaged across participants, within conditions. Black dots show individuals' performance.  Error bars show 95\% confidence intervals. 
		\emph{(B)} Average Dependency over subjects for both open-loop and closed-loop conditions.  Blue bars show Dependency assuming underlying event structure referred to as `data', and red bars are assuming all cue-targ trial types are independent referred to as `indp'. Error bars show 95\% confidence intervals. 
		\emph{(C)} Dependency difference as a function of block.   Dash lines show mean across blocks, and error bars show 95\% confidence intervals.
	}
	\label{fig:deevEEGBeh}
\end{figure}

@@

Building from the theta-phase model \cite{KetzMorkondaOReilly13}, a mechanism supporting perceptual inhibition in service of long-term memory encoding was purposed.  This encoding mechanism in fact works through the successful retrieval of previously encoded information, further strengthening previous associations and binding them to new experiences. This comprehensive model of the hippocampus and surrounding cortex was then used to explore a multi-element memory paradigm shown to differentially engage the hippocampus, referred to as the Dependent Events paradigm.   Here, the mechanisms relating hippocampal reactivation to theta oscillations and attentional filtering to alpha oscillations provided explanatory power in understanding why different association structures lead to more pattern completion, i.e. closed-loop events, as compared to others, i.e. open-loop events. The results showed that repeated reactivation of information incidental to the current task, i.e. in the closed-loop events, strengthen the underlying association structure of those memories, while memory traces which didn't experience this reactivation, i.e.  open-loop events, showed more attentional filtering of cortical processing.  This filtering blocked reactivation of weaker associations, and instead strengthened only the cue-target associations.  The implications of this attentional selection process revealed a potential mechanism to explain the behavioral differences in the Dependent Events paradigm, and more generally provides a model of how retrieval processes can both strengthen are degrade associations depending on task goals as well as memory strength.

>>>>

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information in stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declaritive memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically that oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding a retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the underlying representation that information is encoded into.

@@

It is generally accepted that long-term memory encoding and retrieval are not isolated functions of any one brain area \cite{Gabrieli98,PrinceDaselaarCabeza05}.  However, there are many challenges in moving beyond the traditional studies focusing on individual regions  (e.g., concentrating on the prominent role of the hippocampus), to obtain insight into the nature of network-level dynamics.  An increasing body of work support the idea that oscillatory dynamics in multiple brain regions are associated with different aspects of  memory encoding and retrieval \cite{NyhusCurran10,HanslmayrStaudiglFellner12,Klimesch12}, %update refs here!!!
These studies have generally found a consistent relationship between several frequency bands and successful encoding or retrieval of experiences; namely that oscillatory power within the alpha (8 to 12Hz) and beta (13 to 30Hz) frequency bands generally decrease with these memory processes, while theta (3 to 8Hz) and gamma (30 to 100+Hz) increase in the regions directly involved in the memory operations. Why these relationships exist, and what they tell us about the larger functional organization of memory systems in the brain are still open questions.

One popular model suggests that thalamic circuits play a functional role in selectively gating communication between cortical regions by synchronizing them in various frequency bands \cite{ShermanGuillery06,SaalmannKastner11,KetzJensenOReilly15}.  These gating mechanisms, implemented  by oscillatory synchronization, are central to system level brain function and ultimately affect cognitive behavior such as memory encoding and retrieval.  Further, the particular oscillatory dynamics engaged by this selection process may provide indicators of which system-scale neural circuits are coordinating together to facilitate that particular form of memory processing.

A particular thalmo-cortical circuit of interest for memory processes involves the hippocampus and its associated subcortical network, including the anterior thalamus, and medial septum which are essential in the maintenance of theta oscillations (4 to 8 Hz) \cite{Buzsaki02}.  Theta oscillations have been shown to be critically involved in successful encoding and retrieval of episodic like memories in both humans and animals \cite{NyhusCurran10,JonesWilson05,FuentemillaBarnesDuzelEtAl14}, and it is believed to be primarily driven by the pacemaker cells within the medial septum.  The anterior thalamus, however,  provides the connectivity for the hippocampus to potentially modulate communication with the medial and orbital PFC through theta synchronization \cite{VanderWerfJollesWitterEtAl03,AggletonDumontWarburton11}.  Thus, the theta circuit is important for the core functioning of the hippocampus during encoding and retrieval.

@@

To test this relationship we choose a recent hippocampal dependent multi-item association study to base our computational and empirical investigations on \cite{HornerBisbyBushEtAl15}.  In this study the authors used a previously developed paradigm (see \textcite{HornerBurgess13,HornerBurgess14}) in which they manipulated the binding of a set of overlapping paired associates.   Here two types of associative stimuli were studied, closed-loop and open-loop. In the `closed-loop' events participants studied a set of paired associates on separate trials, e.g. Person-Location, Location-Object, and Object-Person, where each category refers to the same stimulus for a given event.  This created a complete set of binding between all elements.   In contrast, the `open-loop' events incorporated a new element and left some relationships unstudied, e.g. Object-Location, Location-Person, Person-Animal.  Each of the associations was later tested in a cued recognition task, which showed greater inter-event Dependency for the closed-loop compared to the open-loop associates.  Dependency here is operationalized as the retrieval of \emph{all} associated pairs from a given event, or none; whereas no Dependency would imply retrieval of some of the associated pairs but not all.  The goal of this metric is to capture the strength of the binding between elements within a given event, where high Dependency indicates strong associations leading to retrieval of all items within a given event, and low Dependency indicates weak associations leading to some of the items being retrieved incorrectly. In contrast, a non-associative measure of accuracy can be made using an Independent model which calculates Dependency based on the assumption that each retrieval trial is statistically independent from the next, see Section \ref{sec:CalcDep} for more details.  %This provides a nice measure of both Recollection yielding high Dependency from strong associations between elements, but also uses a recognition paradigm where successful retrieval with low between element Dependency implies a similar construct to Familiarity, i.e. retrieval without detailed recall.

@@

The other critical aspect of this bottom-up attention mechanism is the connection to alpha oscillations.  Historically, alpha power has been shown to correlate with the onset of sleep, which led to the suggestion that high alpha power indicates an idle or perhaps inhibited cortical region.  The mechanistic connection between alpha power and thalamic gating is capture in the firing properties of the thalamic relay cells (TRC) which show two distinct modes of firing.  The first shows oscillatory firing patterns acting like a pacemaker system in the alpha frequency range (approximately 10 Hz or 100ms periodicity), and is generally referred to as bursting.  This bursting mode is believed to be engage when a given cortical region is in an idle state, i.e. when no active processing is occurring in the thalamus or the corresponding cortex linked to that thalamic relay region.  The other regime, referred to as tonic firing, shows no specific oscillatory firing pattern, and instead shows increased spontaneous firing in both the thalamus and corresponding cortical regions \cite{LerescheLightowlerSolteszEtAl91,ShermanGuillery06,LopesdaSilva91}.  These dynamics have been studied extensively as a naturally occurring phenomena of thalamic interactions with cortex, however simulation studies have also shown the potential for these mechanisms to be used in a goal directed way \cite{HindriksvanPutten13,VijayanKopell12,LeeWhittingtonKopell13}. The implications of these results are that cortical regions being driving by TRCs in bust mode will show more alpha power, and will process the incoming cortical-cortical information less veridically, while cortical regions being driven by tonic TRC firing will show a relative decrease in alpha power and process cortical-cortical information more completely.  This is the foundation for the connection between alpha power and thalamic gating, such that increased alpha power is indicative of decreased cortical processing.

@@

In this work we start with the Theta-Phase model of the hippocampus and entorhinal cortex, as described in \textcite{KetzMorkondaOReilly13} and implemented in the Emergent framework \cite{AisaMingusOReilly08}.  Briefly, this model is built upon a series of structural and functional hypotheses based on anatomical and physiological data, which have been captured in the complementary learning systems (CLS) model of the hippocampus \cite{McClellandMcNaughtonOReilly95, NormanOReilly03}. The Entorhinal Cortex (EC) in the model is assumed to be the cortical gateway to the hippocampus.  This gateway feeds through the tri-synaptic pathway to the Dentate Gyrus (DG), CA3, and then to CA1.  Similarly, there is a parallel connection through the mono-synaptic pathway from the EC to the CA1 (and back).

@@

The learning dynamics within this model, as governed by the LEABRA algorithm, are dictated by the phases of a presumed theta oscillation, thus its shot-hand name of 'Theta-Phase' used here. In short, during learning two error signals are derived within every period of a give theta oscillation.  The first, during the trough of the theta oscillation, provides an error driven learning signal to the mono-synaptic pathway, learning the invertible mapping between EC and CA1 can be thought of as an encoding.  The second, during the peak of the theta oscillation, provides a error driven learning signal on the tri-synaptic pathway, learning to recollect missing or incomplete information can be thought of as retrieval.

@@

\subsection{Testing Procedure}
\label{sec:TestProc}
The encoding phase was split into three mini-blocks. One pairwise association for each of the 36 events was presented during each mini-block. Presentation order within each mini-block was randomized. For the closed-loop events, the order of presentation across the mini-blocks for a specific event was either: (1) location-object/animal, location-person and person-object/animal, (2) person-object/animal, location-person and location-object/animal, (3) location-person, location-object/ animal, person-object/animal or (4) location-person, person-object/animal and location-object/animal. For open-loop events, it was either: (1) location-object, location-person, person-animal, (2) person-animal, location-person, location-object, (3) location-person, location-object, person-animal or (4) location-person, person-animal, location-object. Each of the four orderings was run on two random initialization of weights, to make up the pool of results. The model was trained on 2 epochs (random permutations of all the trials within a mini-block) for each mini-block before starting training on the next mini-block.  Thus, for the first and second mini-blocks, the closed-loop and open-loop conditions were identical, presenting a single pairwise association and then a second overlapping pairwise association. %The third encoding trial either formed an all-to-all associative structure in the closed-loop condition or an associative chain in the open-loop condition (Fig. 1).

At retrieval, the model was tested on each encoded pairwise association for each event in both directions.  In this way, each event was tested across six separate retrieval trials.  In the behavioral paradigm each retrieval trial was a six-forced choice recognition test with a particular cue element type and potential 6 targets all from the same element type (for example, cue location, retrieve the associated person among five people from other events).  However, in the model, this was captured by providing an activation of 0.1429 to all units in the target element dimension, providing a similar net activation compared to the complete patterns (i.e. $49 * 0.1429=7$), and thus opening the thalamic gate to allow processing in the target dimension. Note that a single cue pattern, and non-informative target dimension activation were presented on all retrieval trials with the task being to reactivate the original paired associate.  This reactivation success was measured by the binary Name Error metric described in Section \ref{sec:HippModelValidation}, i.e. either correct reactivation or not for a given cue-target pair. Thus, as with encoding, the closed-loop and open-loop conditions were exactly matched in terms of stimuli and task demands, differing only in the potential occurrence of incidental reactivation (that is, different levels of pattern completion).  An example of what this looks like in the network is shown in Figure \ref{fig:netNonTarg}B, with more pattern completion occurring in the location stimulus dimension (both in ECout and the TRC layer) for the closed-loop condition compared to open-loop.

@@

The start of each cued recognition trial showed a fixation cross with 0.5 to 1.5 s uniform random duration, followed by the presentation of the cue word above an array of 6 target words (stacked 3 horizontally by 2 vertically, numerically label from left to right and top to bottom), and the fixation cross between them.  The cue word and one of the target words were randomly drawn from a particular event, and the other 5 target words were drawn from the same element type as the correct target word, but randomly from other events.  Locations of the target words in the array were determined randomly.  Participants were instructed to press the numeric key corresponding to the target word that was previously studied with the cue word.  They were not made aware of the underlying event structure, or the open and closed-loop conditions.

@@

All data processing steps and analyses were done in MATLAB using in-house scripts (mat-mvm, n.d.) and the FieldTrip toolbox \cite{OostenveldFriesMarisEtAl11}. A high-pass filter at 0.1 Hz, low-pass filter at 100 Hz, and a notch filter from 59--61 Hz were applied to the data.  ICA artifact removal targeting blinks, and movement artifacts was done on the continuous (i.e. pre-epoched) EEG data after removal of large artifacts through coarse visual inspection.  A semi-automated approach to artifact estimation and detection was done using the SASICA plugin ported for FieldTrip data structures, and methods described in \textcite{ChaumonBishopBusch15}.  After ICA artifact removal, each cued recognition trial was epoched in two ways: stimulus locked (SL) and response locked (RL).  SL epochs were 5000 ms segments, 1 s before the onset of the cue and target array, and 4 s after. RL epochs were also 5000 ms segements, 4 s before target response was made and 1 s after.  Fieldtrips's automatic artifact detection was used to reject particularly noisy epochs as determined by a $z$-score threshold of 20, as well as those that exceed an amplitude of $\pm$200 $\mu$V or a maximum change in amplitude of 200 $\mu$V.

@@

EEG results were analyzed according to the difference between open and closed-loop conditions.  To achieve this, $time \times electrode$ clusters were determined across subjects for each a-priori defined frequency band using a non-parametric statistical test \cite{MarisOostenveld07}.  Frequency data were first averaged over a the defined frequency bands: 3-8 Hz for theta, 8-12 Hz for alpha and 12-30 Hz beta.  These data were then analyzed across all electrodes except the 4 surrounding the eyes, and across the time window of 500 to 3000 ms relative to the cue-target array onset in the stimulus locked epochs, and -3000 ms to -500 ms relative to the button press for a given trial in response locked epochs.  Clustering was done by performing a $t$-test comparing open and closed-loop conditions within each time/electrode bin across subjects.  These $t$-values were then thresholded at $p<.05$, and clustered together based on spatial and temporal adjacency.   Cluster significance was calculated using a monte-carlo style permutation test of the summed $t$-values within a given cluster.  Each observed cluster was subject to 500 random permutations of condition labels where its significance was estimated by the proportion of random permutations which yielded clusters that had a summed $t$-value as large or larger than the observed cluster.

\subsubsection{Representation Similarity Analysis}
\label{sec:deevRSA}
The Representation Similarity Analysis (RSA), was motivated by the non-target reactivation seen in \textcite{HornerBisbyBushEtAl15} and was based on the methods originally developed by \textcite{KriegeskorteMurBandettini08}.  The underlying structure used in the RSA was based on individual retrieval trials within a given event and was done in two ways.  The first was the standard correlation approach where a similarity matrix was constructed by correlating the full $time \times frequency \times electrode$ matrix, splayed out to a single vector, pairwise with each retrieval trial within a given event.  Ideally this results in a 6x6 matrix with all 6 retrieval trials correlated with each other, however some retrieval trials were removed due to the presence of artifacts and therefore all events with more then 3 intact retrieval trials were used in the RSA.  The off diagonal $R$-values in this matrix were then converted to $z$-values using the inverse hyperbolic tangent, and averaged to achieve a single `similarity' measure across retrieval trials for a given event.  These were then averaged across condition with the prediction that more non-target reactivation would manifest in a higher similarity value averaged across events within a given condition.

@@

\begin{figure}
    \begin{center}
	\includegraphics[width=\textwidth]{figs/depNonTarg.pdf}
    \end{center}
      \caption[Non Target Reactivation]{
       \emph{(A)} Name Error accuracy for non-target reactivation in open-loop and closed-loop events.  The horizontal axis labels correspond to the cue and target respectively, and the non-target element is the one not listed in the label.  For example `loc-per' indicates a location cue and person target trial, and therefore the non-target element is either object or animal depending on the encoding paris for the given event.  The 'oth' label corresponds to either the object or animal element again depending on the corresponding encoding pairs for the given event.  Note, non-target accuracy was not assessed for object or animal elements as cues, as open-loop events have no corresponding non-target element in these cases, and would be a biased comparison with closed-loop events which do. Black dots indicate individual network runs with different training orders (see section \ref{sec:TestProc}), red bars are closed-loop and blue bars are open-loop, dash lines are means across all cue-targ trial types, and error bars are 95\% confidence intervals. \emph{(B)} Example cued retrieval trials from open-loop (top) and closed-loop (bottom) events.  Both show a retrieval trial with a person cue (unit group 2, numbering from lower left proceeding up), and a animal target (unit group 4), and a location non-target (unit group 1).  TRC shows bottom-up (person and animal unit groups) as well as hippocampal driven (Location unit group) modulation of thalamic gate.
	}
     \label{fig:netNonTarg}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{figs/deevEEGBeh.pdf}
	\end{center}
	\caption[EEG Dependent Events Behavior]{
		\emph{(A)} Cued recognition accuracy for all cue-targ and targ-cue combinations, averaged across participants, within conditions. Black dots show individuals' performance.  Error bars show 95\% confidence intervals. 
		\emph{(B)} Relationship between retrieval confidence and accuracy.  Scatter plot of average cued-recognition accuracy within subjects as a function of self rated confidence, broken down by open and closed-loop conditions.  Overlayed histograms show smoothed marginal distributions for open and closed-loop conditions.
		\emph{(C)} Dependency difference as a function of block. Dependency assuming underlying event structure referred to as `data', and assuming all cue-targ trial types are independent referred to as `Independent'.  Dash lines show mean across blocks, and error bars show 95\% confidence intervals.
	}
	\label{fig:deevEEGBeh}
\end{figure}

@@

Building from the theta-phase model \cite{KetzMorkondaOReilly13}, a mechanism supporting perceptual inhibition in service of long-term memory encoding was purposed.  This comprehensive model of the hippocampus and surrounding cortex was then used to explore a multi-element memory paradigm shown to differentially engage the hippocampus, referred to as the Dependent Events paradigm.   Here, the mechanisms relating hippocampal reactivation to theta oscillations and attentional filtering to alpha oscillations provided explanatory power in understanding why different association structures lead to more pattern completion, i.e. closed-loop events, as compared to others, i.e. open-loop events. The results showed that repeated reactivation of information incidental to the current task, i.e. in the closed-loop events, strengthen the underlying association structure of those memories, while memory traces which didn't experience this reactivation, i.e.  open-loop events, showed more attentional filtering of cortical processing.  This filtering blocked reactivation of weaker associations, and instead strengthened only the cue-target associations.  The implications of this attentional selection process revealed a potential mechanism to explain the behavioral differences in the Dependent Events paradigm, and more generally provides a model of how retrieval processes can both strengthen are degrade associations depending on task goals as well as memory strength.
